---
title: "2012 NLS Data RMD"
author: "Scott Stoner"
date: '2023'
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(dplyr)
library(ggplot2)
```

## Data Import

This section covers importing lake chemistry and watershed data for the purpose of lake selection. The relevant data is first retrieved from one of the NLS 2012 csv files. Then each table is reduced to contain just the variables of interest. This table is converted into a tibble, which overwrites the original variable containing the imported data. These variables fall into two categories, watershed and water chemistry. Additional variables are included to help with ID or shed light on connections made during PCA. Important to note that there are 2 ID systems in use, UIDs (unique sample event IDs) and Site IDs (lake specific IDs). Retaining both is important to select and summarize lakes with multiple sample events later in the process. Variable names and notes are included for each imported data set. \
\

#### Site Info

- AREA_HA (lake area, hectares)
- STATE
- SIZE_CLASS
- SITETYPE (probability or handpicked)
- ELEVATION
- FS_EW (forest service land east or west)
- DSGN12 (incorporated in the NLA2012 design framework)
- NA_L2CODE (ecoregion level 2)
- NA_L3CODE (ecoregion level 3)
- URBAN (determination based on land use)
- EVAL_NAME (lake name)
- LAKE_ORIGIN (“manmade or natural”)
- PERIM_KM (lake perimeter)
- SITESAMP (was site sampled)
- DATE_COL (date of sample collection)
- SITE_ID (id code for site 2012 version)
- VISIT_NO (visit number for site)
- UID (unique site visit id)

``` {r import_siteinfo}
siteinfo <- read.csv("nla2012_wide_siteinfo_08232016.csv")

siteinfo %>% 
  dplyr::select(UID, SITE_ID, VISIT_NO, DATE_COL, SITETYPE, SITESAMP, DSGN12, EVAL_NAME, STATE, ELEVATION, AREA_HA, PERIM_KM, SIZE_CLASS, FS_EW, NA_L2CODE, NA_L3CODE, URBAN, LAKE_ORIGIN) %>%
  tibble::as_tibble() -> siteinfo
  
```

#### Watershed Data

- SITE_ID
- DOMGEOL_BSN (dominant geology)
- NLCD2006_ASSIGNEDAREA_BSN (area within basin assigned nlcd value*)
- NLCD2006_DEVELOPEDPCT_BSN (percent developed)
- NLCD2006_FORESTPCT_BSN (percent forest)
- NLCD2006_IMPERVAREA_BSN (percent impervious)
- PMEAN_BSN (mean annual precipitation in basin)[multiple precip values]
- TMEANPY_BSN (mean temperature in basin 12 months)[multiple temp values]


``` {r import_watershed}
watershed <- read.csv("nla2012_wide_watershed.csv")

watershed %>%
  dplyr::select(SITE_ID, DOMGEOL_BSN, NLCD2006_ASSIGNEDAREA_BSN, NLCD2006_DEVELOPEDPCT_BSN, NLCD2006_FORESTPCT_BSN, NLCD2006_IMPERVAREA_BSN, PMEAN_BSN, TMEANPY_BSN) %>%
  tibble::as_tibble() -> watershed
```

#### Water Chemistry Data

- UID
- NTL_RESULT (Total Nitrogen mg/L)
- PTL_RESULT (Total Phosphorus ug/L)
- PH_RESULT

``` {r import_waterchem}
waterchem <- read.csv("nla2012_waterchem_wide.csv")

waterchem %>%
  dplyr::select(UID, NTL_RESULT, PTL_RESULT, PH_RESULT) %>%
  tibble::as_tibble() -> waterchem
```

## Data Merge

The selected data from the imported tables is now combined into a single table that will be used for lake selection. First a left join using the SITE_ID field to append watershed data to general site info. Then a left join is used to append the water chemistry data to each visit using the UID field. The result is output into the 'mega_table'.

``` {r mega_merge}
siteinfo %>%
  dplyr::left_join(watershed, by = "SITE_ID") %>%
  dplyr::left_join(waterchem, by = "UID") -> mega_table
```

### Lake Selection \

#### Initial Filter

This section concerns removing data (lakes) that lack sampling data. First, lakes with no sampling data are excluded, followed by lakes that were not randomly selected.

``` {r initial_filter}
mega_table %>%
  dplyr::filter(SITESAMP == "Y") %>%
  dplyr::filter(SITETYPE == "PROB") -> initial_filter_result
```

#### Multiple Sample Lakes

This section currently strips the second sample set from each lake with multiple samples.

``` {r second_sample_strip}
initial_filter_result %>%
  dplyr::filter(VISIT_NO == 1) -> data_filtered
```


#### Watershed Filter

Here is where the bulk of the lake selection process happens. 

List of selection conditions in order:

1. NA_L2CODE = 6.2
  - The 6.2 NA Ecoregion 2 Code represents the "Western Cordillera" also known as the Northwestern Forested Mountain Region.
2. STATE = CA, OR, WA, ID, MT
  - The restriction to these 5 states removes areas of Ecoregion 6.2 that are mostly in Colorado and New Mexico.
3. Possibly more later: Elevation, Size, Land Use

``` {r lake_selection}
data_filtered %>%
  dplyr::filter(NA_L2CODE == 6.2) %>%
  dplyr::filter(STATE == "CA" | STATE == "OR" | STATE == "WA" | STATE == "ID" | STATE == "MT") -> data_filtered
  data_filtered
```

## Data Summary

These are summary statistics for the variables of interest. These are set up to exclude variables that have already been controlled for (e.g. Ecoregion, SITESAMP etc.). Pushing this into the data_summarized object to allow for later sections to use the whole table. **This section incomplete:** working on utilizing the dplyr:summarise() function. 

``` {r data_summary}
data_filtered %>%
  dplyr::select(SITE_ID, AREA_HA, SIZE_CLASS, ELEVATION, NA_L3CODE, URBAN, LAKE_ORIGIN, DOMGEOL_BSN, NLCD2006_ASSIGNEDAREA_BSN,NLCD2006_DEVELOPEDPCT_BSN, NLCD2006_FORESTPCT_BSN,NLCD2006_IMPERVAREA_BSN, NLCD2006_FORESTPCT_BSN,
NLCD2006_IMPERVAREA_BSN, PMEAN_BSN, TMEANPY_BSN, NTL_RESULT, PTL_RESULT, PH_RESULT) -> data_summarized
```

#### Water Quality

This section includes graphical representations of water quality parameters.

``` {r water_quality_scatterplots}
ggplot(data_filtered, aes(x = NTL_RESULT, y = PTL_RESULT)) +
  geom_point() +
  geom_smooth(method='lm', se=FALSE) +
  scale_x_continuous(trans='log') +
  scale_y_continuous(trans='log')

ggplot(data_filtered, aes(x = PH_RESULT, y = NTL_RESULT)) +
  geom_point() +
  geom_smooth(method='lm', se=FALSE)

ggplot(data_filtered, aes(x = PH_RESULT, y = PTL_RESULT)) +
  geom_point() +
  geom_smooth(method='lm', se=FALSE)
```
\
This is a linear regression for phosphorus~nitrogen and phosphorus~pH.

```{r water_quality_linreg}
summary(lm(log(PTL_RESULT)~log(NTL_RESULT), data = data_filtered))

summary(lm(PTL_RESULT~PH_RESULT, data = data_filtered))
```

``` {r water_quality_boxplots}
ggplot(data_filtered, aes(x=as.factor(SITESAMP), y=NTL_RESULT)) +
  geom_boxplot()  +
  labs(x="all sites", title = "nitrogen")

ggplot(data_filtered, aes(x=as.factor(SITESAMP), y=PTL_RESULT)) +
  geom_boxplot()  +
  labs(x="all sites", title = "phosphorus")

ggplot(data_filtered, aes(x=as.factor(SITESAMP), y=NTL_RESULT)) +
  geom_boxplot()  +
  labs(x="all sites", title = "nitrogen - log transformed") +
  scale_y_continuous(trans='log10')

ggplot(data_filtered, aes(x=as.factor(SITESAMP), y=PTL_RESULT)) +
  geom_boxplot()  +
  labs(x="all sites", title = "phosphorus - log transformed") + 
  scale_y_continuous(trans='log10')

ggplot(data_filtered, aes(x=as.factor(SITESAMP), y=PH_RESULT)) +
  geom_boxplot()  +
  labs(x="all sites", title = "ph")
```

``` {r water_quality_summary}
summary(data_filtered)
```

#### Watershed Characteristics

This section includes graphical representations of watershed characteristics.

### Establish Gradient

This section sorts data based on the variable of interest: phosphorus.

``` {r gradient_establisher}
data_filtered %>%
  dplyr::arrange(PTL_RESULT) -> data_filtered_gradient
  data_filtered_gradient
```

### PCA 

This section is where the PCA is conducted on lake chemistry data.

``` {r PCA}
data_filtered %>%
  dplyr::select(PTL_RESULT, NTL_RESULT, PH_RESULT) %>%
  svd() -> pca_result

pca_result$v
pca_result$d
```