---
title: "2012 NLS Data RMD"
author: "Scott Stoner"
date: "2023"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Data Import and Merge

This section covers importing lake chemistry and watershed data for the purpose of lake selection.

The relevant data is retrieved from several of the NLS 2012 csv files.

```{r data-import}
siteinfo<-read.csv("nla2012_wide_siteinfo_08232016.csv")
watershed<-read.csv("nla2012_wide_watershed.csv")
keyvariables<-read.csv("nla12_keyvariables_data.csv")
waterchem<-read.csv("nla2012_waterchem_wide.csv")
```

Each table is converted into a tibble for ease of manipulation.

```{r data-to-tibble}
siteinfo<- tibble::as_tibble(siteinfo)
watershed<- tibble::as_tibble(watershed)
keyvariables<- tibble::as_tibble(keyvariables)
waterchem<- tibble::as_tibble(waterchem)
```

Each table needs to be pruned to contain just the variables of interest. Starting with the largest (by row) table, siteinfo. The current list of variables of interest with notes: 

- AREA_HA (lake area, hectares)
- STATE
- SIZE_CLASS
- SITETYPE (probability or handpicked)
- ELEVATION
- FS_EW (forest service land east or west)
- DSGN12 (incorporated in the NLA2012 design framework)
- NA_L2CODE (ecoregion level 2)
- NA_L3CODE (ecoregion level 3)
- URBAN (determination based on land use)
- EVAL_NAME (lake name)
- LAKE_ORIGIN (“manmade or natural”)
- PERIM_KM (lake perimeter)
- SITESAMP (was site sampled)
- DATE_COL (date of sample collection)
- SITE_ID (id code for site 2012 version)
- VISIT_NO (visit number for site)
- UID (unique site visit id)

```{r data-consolidation}
siteinfo_select<- dplyr::select(siteinfo, UID, SITE_ID, VISIT_NO, DATE_COL, SITETYPE, SITESAMP, DSGN12, EVAL_NAME, STATE, ELEVATION, AREA_HA, PERIM_KM, SIZE_CLASS, FS_EW, NA_L2CODE, NA_L3CODE, URBAN, LAKE_ORIGIN)
```

The other data tables need to be pruned as well. 

The first table to be appended is 'watershed', which will contain these variables:

- SITE_ID
- DOMGEOL_BSN (dominant geology)
- ELEV_SITE (second elevation method)
- ELEVMEAN_BSN (mean basin elevation)
- NLCD2006_ASSIGNEDAREA_BSN (area within basin assigned nlcd value*)
- NLCD2006_DEVELOPEDPCT_BSN (percent developed)
- NLCD2006_FORESTPCT_BSN (percent forest)
- NLCD2006_IMPERVAREA_BSN (percent impervious)
- PMEAN_BSN (mean annual precipitation in basin)[multiple precip values]
- TMEANPY_BSN (mean temperature in basin 12 months)[multiple temp values]

``` {r watershed-chop}
watershed_select<- dplyr::select(watershed, SITE_ID, DOMGEOL_BSN, ELEV_SITE, ELEVMEAN_BSN, NLCD2006_ASSIGNEDAREA_BSN, NLCD2006_DEVELOPEDPCT_BSN, NLCD2006_FORESTPCT_BSN, NLCD2006_IMPERVAREA_BSN, PMEAN_BSN, TMEANPY_BSN)
```

Next is the data in keyvariables:

- UID
- CHLX_RESULT (chlorophyll a)
- NTL_RESULT (Total Nitrogen mg/L)
- PTL_RESULT (Total Phosphorus ug/L)

``` {r keyvariables-chop}
keyvariables_select<- dplyr::select(keyvariables, UID, CHLX_RESULT, NTL_RESULT, PTL_RESULT)
```

Last is the water chemistry data:

- UID
- PH_RESULT

``` {r waterchem-chop}
waterchem_select<- dplyr::select(waterchem, UID, PH_RESULT)
```

Now these tables all need to be combined into the mega_table, which is the output of this section. First a left join using the SITE_ID field to append watershed data to general site info.

``` {r mega_merge1}
mega_table<- dplyr::left_join(siteinfo_select, watershed_select, by = "SITE_ID")
```

Now both sets of water chemistry data need to be added, first keyvariables data.

``` {r mega_merge2}
mega_table<- dplyr::left_join(mega_table, keyvariables_select, by = "UID")
```

Then waterchem data.

``` {r mega_merge3}
mega_table<- dplyr::left_join(mega_table, waterchem_select, by = "UID")
```

This is the resulting fully merged table. It still contains a lot of unwanted data that will need to be removed before analysis.

``` {r merge_result, echo=FALSE}
mega_table
```

## Initial Data Filter

This section concerns removing data (lakes) that lack sampling data. There will also be a component that handles sites with multiple samples.

First, the initial_filter data table is created. This is the output for this section and a potential point of comparison for future refined lake data sets.

``` {r initial_filter_created}
initial_filter<- mega_table
```

Lakes with no sampling data are removed.

``` {r missingsample_filter}
initial_filter<- dplyr::filter(initial_filter, SITESAMP == "Y")
```

Lakes that were hand-selected are removed to reduce bias and due to lack of water chemistry data.

``` {r handselected_filter}
initial_filter<- dplyr::filter(initial_filter, SITETYPE == "PROB")
```

IN PROGRESS: Multiple Sample Lakes

The resulting data set lacks any critical missing values and is ready to be used for lake selection.

``` {r initial_filter_result, echo=FALSE}
initial_filter
```
## Data Filter

This section contains the lake selection process and the reasoning behind those decisions.